Reprotool: a debugging tool to identify numerical errors and assess reproducibility of pipelines

### 2019-03-19 paper.tex

\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
\newcommand{\reprozip}[0]{ReproZip}

\title{Numerical error propagation in the HCP structural pre-processing 
pipelines}
\title{Reprotool: a debugging tool to identify numerical errors 
and assess reproducibility of the neuroimaging pipelines}

@@

Many experiments show that computational analysis results are still not 
completely reproducible. Although new technologies such as 
virtualization , version control systems , and provenance management 
tools provide more reliable environment and significantly improve the 
reproducibility crisis. 
In particular, computational environments are known to have an effect 
on the results produced by neuroimaging pipelines, presumably due to 
the creation, propagation and amplification of small numerical errors 
across the pipelines. Similarly, small perturbations that originated 
from computational environments including diverse operating systems, 
hardware categories, software version, etc. highlight the numerical 
errors.
However, the precise causes of such instabilities and the path along 
which they propagate in the pipelines are unclear.  We present a 
technique to identify the processes in the pipeline that create numerical 
errors along the execution, and we apply this technique to the HCP 
structural pre-processing pipelines.

@@

Research findings are expected to be reproducible to evaluate the 
authenticity and reliability of the claims.There are different 
terminologies for the terms reproducibility, repeatability, and 
replicability over the time. For example, the Association for Computing 
Machinery (ACM)~\cite{acm2016terminology} proposes three different 
definitions. From the ACM's points of view, repeatability is defined as 
repeating the computation by the same experimental setup including 
operator team, operating conditions, location, and measuring system. 
Replicability has almost the similar definition with repeatability 
which use the identical experimental conditions except performer team 
which means that an independent group can achieve the same results. 
Beside, reproducibility is measured by performing different 
experimental setup via different laboratory team independently. In this 
paper, we follow Peng's~\cite{peng2011reproducible} definition of 
reproducibility which is assessed by performing the experiment using 
the same data and the same analytic methods and expecting the same 
results. A reproducible study provides a context in which, one is able 
to get the consistence results to the original work. In addition, it 
enables others to use the existing work as a part of their 
experiments~\cite{plesser2018reproducibility}. It not only saves a 
great deal of time, but also contributes to focusing on the other 
parts.

Reproducibility of computational pipelines has been repeatedly 
investigated so far. The comparisons show that the analysis results are 
not completely reproducible across the small perturbations which have 
emerged by changing the execution environments including operating 
system, software and analysis method, and hardware configuration. There 
are different studies that discovered factors that introduce such tiny 
perturbations and consequently hamper the reproducibility of the 
experiments. In particular, Groenschild \emph{et al.} first identified 
the effect of operating systems on Freesurfer 
results~\cite{Gronenschild2012}. Similarly, the effect of OS is 
quantified in~\cite{Glatard2015} more specifically on some 
of the main neuroimaging pipelines including several FSL pipelines, 
CIVET and Freesurfer. The results of comparisons indicate a significant 
errors between analysis results of the same experiment on different 
operating system.

It should be noted that \emph{error} is defined as the statistical 
differences between results of two experiments in this work. Moreover, 
the operating system is defined here as a consistent set of software 
packages organized in a trusted repository.

The operating system is not the only part of the computational 
environment that may hamper reproducibility. Many other computational 
factors including hardware configuration, software version, and 
parallelization techniques are reported~\cite{diethelm2012limits} as 
the main influential computing infrastructure on reproducibility.

It has revealed that the majority of analysis results are not 
numerically reproducible because of arithmetic operations which is 
associated with the type of execution such as sequential or parallel 
executions, floating-point operations, different standard calculations 
like IEEE-754 or other standards, etc.

In order to overcome these issues and making reproducible analysis, a 
comprehensive solution requires which not only encapsulates and 
re-executes all the software/hardware dependencies but also fixes the 
numerical instabilities. Therefore, we introduce a framework to 
identify the processes in the pipeline which are responsible for such 
numerical errors along the execution. In this paper, the execution 
results of the pipelines across different operating systems are 
considered to find out the cause of the reproducibility issues using an 
interposition techniques.

Figure~\ref{fig:overview} illustrate an overview of the proposed 
technique. First, we execute pipelines in a consistence environment 
using docker container and Boutiques framework. In the next step, the 
origin of errors, processes that make differences, are detected through 
a classification approach. Processes are classified by several 
modifications iteratively and finally, it creates corresponding process 
graph of the pipeline execution. We will explain each step in detail in 
the following sections.

\begin{figure}[H]
\centering
  \includegraphics[scale=0.5]{images/overview.png}
  \caption{An overview of proposed technique.}
  \label{fig:overview}
\end{figure}

\paragraph{} The reminder of this paper explains the proposed debugging tool in two 
parts. First, the execution environment of the pipelines and then the 
classification procedure of the processes are clarified. We describe 
the HCP preprocessing pipelines as a popular neuroimaging pipelines to 
test and apply reprotool. Finally, we report the analytic process which 
are responsible for the errors in the pipelines.

@@

A different process graph may be produced for every subject in every 
execution condition. Among our dataset, we identified 4 types of 
subjects with different numbers of T1w and T2w images. We verified that 
the process graphs were identical for all subjects of the same type, 
for all execution conditions \todo{Check that}. The remainder of the 
analysis is done separately for each subject type.

@@

\paragraph{} In the next step, we classify the processes into three 
categories based on the process graph and the error matrix file 
mentioned in the previous Section:
\begin{enumerate}
\item Processes that read files that do not have errors and write files that do not 
have errors are \emph{transparent} (Figure~\ref{fig:processes}.a).
\item Processes that read files 
that do not have any error but write files that have errors 
\emph{create} errors in the pipeline (Figure~\ref{fig:processes}.b).
\item Processes that read files 
that have errors and write files that do not have errors \emph{remove} 
errors from the pipeline (Figure~\ref{fig:processes}.c).
\item Processes that read files that have errors and write files that also have errors are 
\emph{unknown} (Figure~\ref{fig:processes}.d).
\end{enumerate}

@@

\paragraph{Challenge3:} Another challenge is to classify processes that 
read files that have errors and write files that also have errors. In 
such situations, it is not possible to determine from the pipeline 
results whether the process created errors, removed some errors, or was 
transparent. We label such processes \emph{unknown}. To address this 
issue, we developed an iterative approach that consists of the 
following steps:

@@

The three steps in Figure~\ref{fig:iterations} correspond to the 
iterations of the classification algorithm At step one \texttt{2\#bet2} 
is classified as error creator (red) as it produced files with errors 
from files without error. \texttt{3\#fslstats} is classified as error 
remover (blue) as it produced files without errors from files with 
errors. \texttt{4\#fslmaths} and \texttt{5\#fslstats} are classified as 
unknown (yellow) as they produce files with errors from files with 
errors.

@@

\paragraph{Structural Pre-processing Pipelines} The structural 
pre-processing consists of PreFreeSurfer, FreeSurfer and 
PostFreeSurfer~\cite{glasser2013minimal}. In this paper, we analyzed two 
pipelines, PreFreeSurfer and FreeSurfer.

PreFreeSurfer consist of various steps including correction of MR 
gradient nonlinearity distortions, align the T1w and T2w images using 
FSL FLIRT, Align native space to MNI template using ACPC and FSL FLIRT, 
brain extraction usin FSL FLIRT and FNIRT to MNI template, register T2w 
to T1w using FLIRT's BBR, perform a bias field correction, and register 
the subject's native structural space to MNI space.

@@

\paragraph{} Two types of errors can occur in the subjects due to the 
errors in the operating systems. One is inter-OS error caused by the 
operating system library updates and the other type, intra-OS errors 
occurs as a result of the pseudo-random processes used in the 
pipelines. An example of a pseudo-random process function is, a random 
number generator that would get initialized using a seed state. the 
proposed method can be used to identify both kind of errors. The files 
that are common to all the subjects only are taken into consideration 
for comparison. The first step is identification of files with 
differences in their checksums. This is identified using the checksums 
that are recorded after the processing. Intra-OS errors are identified 
using the run-number added as the suffix for the conditions. For 
example, the two batches of subjects processed under the same condition 
(CentOS6) are stored as run-1 and run-2. The files belonging to the 
subjects stored under the above mentioned conditions are treated as 
intra-OS runs.

\subsection{Inter-OS differences}

\paragraph{Binary differences}

In a previous study~\cite{Scaria2017}, we showed that pre-processing 
pipelines of the Human Connectome Project~\cite{Glasser2013} were 
sensitive to operating system variation so that Figure 
\ref{fig:tissue_class} illustrates the binarized errors after brain tissue 
segmentation process in PreFreeSurfer. In addition, 
Figure~\ref{fig:fnirt_result} shows the same errors occurred in 
non-linear registration process between CentOS6 and CentOS7. 
Figure~\ref{fig:pipeline_errors}.a.

\begin{figure}[H]
\centering
  \includegraphics[scale=0.6]{images/fnirt_result.png} 
  \caption{Absolute errors between FNIRT results from PreFreeSurfer 
  (T2w\_acpc\_to\_MNI\_nonli.nii.gz), subject 104820 (centos6 vs. 
  centos7)~\cite{Scaria2017}. } 
  \label{fig:fnirt_result}
\end{figure}

@@

@@

\paragraph{PreFreeSurfer.} Among the 117 data files produced by 
PreFreeSurfer, 21 did not have any error for any subject, 92 had errors 
for all subjects and 4 had errors for 3 subjects only.

@@

\paragraph{FreeSurfer.} In the second analysis, the result of 
PreFreeSurfer is considered as the input files of FreeSurfer pipeline. 
From the FreeSurfer results, total 61 files were found to be different 
between CentOS6 and CentOS7. The processes that introduce errors in 
FreeSurfer are shown on Figure~\ref{fig:fs_error_table}. Most frequent 
processes that create errors are \emph{mris\_anatomical\_stats} which 
measures a number of anatomical properties, \emph{mris\_make\_surface} 
to generate pial surfaces, first pass, no T2 adjustment, 
\emph{mri\_convert} which normalize T1w and T2w images for the benefit 
of \emph{mris\_make\_surfaces}, and \emph{fslmaths} create and apply a 
binary mask to filling of holes to remove  any non-grey/white tissue 
using a threshold created by FSLSTATS at Mean. In addition, all the 
other processes that created errors in the pipeline at least once are 
categorized as others such as \emph{mri\_ca\_register}, 
\emph{mri\_ca\_normalize}, and \emph{mri\_watershed}. Furthermore, 
processes that remove errors in FreeSurfer represented on 
Figure~\ref{fig:fs_remove_table} including \emph{mri\_convert} and 
\emph{mri\_surf2surf}.

@@

pipeline amplify small numerical differences, they are numerically 
unstable. Furthermore, math libraries evolve over time, leading to 
different numerical errors. we listed some of the irreproducibility 
causes of the pipelines as we mentioned in the previous section 
overally along with exact command arguments.

Mention that this was only possible because the unprocessed data was shared in the first place.
DICOM to Nifti conversion was out of scope and may introduce other issues.

@@

Our thechnique is able to characterize the stability of a pipeline's components automatically.
The numerical instability in the PreFreesurfer HCP pipeline arises mainly from
linear and non-linear registration processes implemented in FSL FLIRT and FNIRT.

@@

The results still suffer from small pertubations literally because of the fact that pipeline are not numerically stable.
The preferred solution is to detect and fix numerical instability of the pipeline instead of masking the problem.
These processes need to be reviewed to understand and correct the cause of instabilities. 
In this correction process, accuracy has to be considered in addition to stability.

\item In the future works we aim to measure the impact of such errors
on the specific domain of the neuroscience and how show the significant it is.

\item The purpose of the project is generalizing pipelines to get a reproducible execution 
on every operating system instead of choosing one main OS to.

>>>>

\newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
\newcommand{\note}[2]{\color{blue}Note: #1\color{black}}

\newcommand{\reprozip}[0]{ReproZip}

\title{Reprotool: a debugging tool to identify numerical errors 
and assess reproducibility of pipelines}

@@

Many experiments show that computational analysis results are still not 
completely reproducible. In particular, computational environments 
including different operating systems, hardware categories, and 
software versions are known to have effects on the results produced by 
analysis pipelines. These effects are presumably due to the creation, 
propagation and amplification of small numerical errors across the 
pipelines. 
Although new techniques such as virtualization , version 
control systems , and provenance management tools provide more reliable 
environment and significantly improve the reproducibility of scientific 
findings. 
However, the precise causes of such instabilities and the path along 
which they propagate in the pipelines are unclear.  We present a 
technique to identify the processes in the pipeline that create 
numerical errors along the execution, and we apply this technique to 
the HCP structural pre-processing pipelines.

@@

Research findings are expected to be reproducible so that 
their authenticity and reliability can be evaluated. 
There are different definitions for the terms reproducibility, 
and replicability, which leads to confusion since the 
same words are used for different concepts.

According to Peng's definition~\cite{peng2011reproducible}, 
reproducibility is defined as the ability to regenerate the same 
results as the original findings when the experiment is reanalyzed 
given exactly the same analytic methods and data. Reproducibility 
ensures that independent scientists can reproduce the same results 
using the same data and procedure as published in the original 
publication. On the contrary, replicability is defined as the ability 
to obtain similar results as published in the original study when 
the experiment is reimplemented using independent data and analytic 
methods. Replicability confirms scientific claims and ensures that 
independent investigators can produce consistent results, using new 
data and methods.

In addition, numerical reproducibility is defined as the ability to 
regenerate bit for bit identical results from multiple 
runs~\cite{hill2017numerical}. Two files will be considered numerically 
reproducible if they have identical binary contents. Binary 
comparison is calculated by comparing checksums. It must be pointed out 
that a computation might be reproducible based on Peng's definition, 
but not be numerically reproducible. For instance, small numerical 
errors created during the pipeline execution may hamper numerical 
reproducibility, but be negligible in the final results.

Reproducibility, as the cornerstone of scientific research, guarantees 
the reliability of results. A reproducible study provides a context in 
which one is able to get results that are consistent with the original 
work. In addition, it not only saves a great deal of time, but also 
enables others to use existing works as a part of their 
experiments~\cite{plesser2018reproducibility}. 
In this paper, we follow Peng's definition of reproducibility 
unless we directly refer to numerical reproducibility. 
We seek to identify the reasons why such reproducibility may not be 
ensured, focusing in particular on computational aspects.

Recently, scientists began to realize that the results of many 
scientific experiments were neither replicable nor reproducible. 
Results show that the variety of computational infrastructures 
including workstation types, parallelization methods, operating 
systems, and analysis packages are known to influence reproducibility 
because of the creation of small numerical 
errors~\cite{Gronenschild2012, diethelm2012limits, Glatard2015}. The 
propagation and amplification of these tiny errors by analysis 
pipelines may cause reproducibility issues. 
\note{In this case, the analysis pipelines are said to be numerically 
unstable. Numerical instability is a characteristic of the pipelines 
which amplify small numerical errors and then hamper the 
reproducibility of the analyses depending on the length of the pipeline 
and magnitude of the errors. In many cases, numerical instability is an 
important issue for reproducibility. }
In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} more specifically on some of 
the main neuroimaging pipelines including FSL, CIVET and Freesurfer 
pipelines. The results of comparisons indicate a significant errors 
between analysis results of the same experiment on different operating 
system. It should be noted that \emph{error} is defined as the binary 
differences between results of two analyses in this paper, which is 
calculated by the checksum method.

\note{The operating system is not the only part of the computational 
environment that may hamper reproducibility. Many other computational 
factors including hardware configuration, software version, and 
parallelization techniques are reported as 
the main influential computing infrastructure on reproducibility. }

There are different solutions to make reproducible analysis including 
containerization techniques to encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities. For this purpose, we introduce a framework to identify 
the processes in pipeline that create numerical errors across different 
runs. In this paper, the execution results of pipelines across 
different operating systems are considered to find out the cause of the 
irreproducibility using an interposition techniques.

Figure~\ref{fig:overview} illustrates an overview of the proposed 
technique. First, we execute pipelines in a consistence environment 
using docker container and Boutiques framework. In the next step, the 
processes that create differences are classified using an iterative 
approach. Finally, we make corresponding process graph of the 
pipeline execution that shows the processes related to the creation of 
errors. We will explain each step in detail in the following sections.

\begin{figure}[H]
\centering
  \includegraphics[scale=0.5]{images/overview.png}
  \caption{An overview of the proposed technique.}
  \label{fig:overview}
\end{figure}

The reminder of this paper explains the Reprotool in different parts as 
the execution environment of the pipelines, methods to measure 
differences, and how to characterize errors. We describe the HCP 
preprocessing pipelines as a popular neuroimaging pipelines to test and 
apply the Reprotool on. Finally, we report the processes that are 
responsible for the errors in the pipelines.

@@

A different process graph may be produced for every subject in every 
execution condition. Among our dataset, we identified 4 types of 
subjects with different numbers of T1w and T2w images. We verified that 
the process graphs were identical for all subjects of the same type, 
for all execution conditions. The remainder of the 
analysis is done separately for each subject type.

@@

In the next step, we classify the processes into three 
categories based on the process graph and the error matrix file 
mentioned in the previous Section:
\begin{enumerate}
\item Processes that read files that do not have errors and write files that do not 
have errors are \emph{transparent} (Figure~\ref{fig:processes}.a).
\item Processes that read files 
that do not have any error but write files that have errors 
\emph{create} errors in the pipeline (Figure~\ref{fig:processes}.b).
\item Processes that read files 
that have errors and write files that do not have errors \emph{remove} 
errors from the pipeline (Figure~\ref{fig:processes}.c).
\item Processes that read files that have errors and write files that also have errors are 
\emph{unknown} (Figure~\ref{fig:processes}.d).
\end{enumerate}

@@

\paragraph{Challenge3:} Another challenge is to classify processes that 
read files that have errors and write files that also have errors. In 
such situations, it is not possible to determine from the pipeline 
results whether the process created errors, removed some errors, or was 
transparent. We label such processes \emph{unknown}.

To address Challenge 3, we developed an iterative approach that 
consists of the following steps:

@@

The three steps in Figure~\ref{fig:iterations} correspond to the 
iterations of the classification algorithm. At step one, \texttt{2\#bet2} 
is classified as error creator (red) as it produced files with errors 
from files without error. \texttt{3\#fslstats} is classified as error 
remover (blue) as it produced files without errors from files with 
errors. \texttt{4\#fslmaths} and \texttt{5\#fslstats} are classified as 
unknown (yellow) as they produce files with errors from files with 
errors.

@@

The structural pre-processing pipelines consists of PreFreeSurfer, 
FreeSurfer and PostFreeSurfer. In this paper, 
we analyzed two pipelines, PreFreeSurfer and FreeSurfer. 
PreFreeSurfer consist of various steps including correction of MR 
gradient nonlinearity distortions, align the T1w and T2w images using 
FSL FLIRT, Align native space to MNI template using ACPC and FSL FLIRT, 
brain extraction usin FSL FLIRT and FNIRT to MNI template, register T2w 
to T1w using FLIRT's BBR, perform a bias field correction, and register 
the subject's native structural space to MNI space.

@@

Two types of errors can occur in the subjects due to the 
errors in the operating systems. One is inter-OS error caused by the 
operating system library updates and the other type, intra-OS errors 
occurs as a result of the pseudo-random processes used in the 
pipelines. An example of a pseudo-random process function is, a random 
number generator that would get initialized using a seed state. the 
proposed method can be used to identify both kind of errors. The files 
that are common to all the subjects only are taken into consideration 
for comparison. The first step is identification of files with 
differences in their checksums. This is identified using the checksums 
that are recorded after the processing. Intra-OS errors are identified 
using the run-number added as the suffix for the conditions. For 
example, the two batches of subjects processed under the same condition 
(CentOS6) are stored as run-1 and run-2. The files belonging to the 
subjects stored under the above mentioned conditions are treated as 
intra-OS runs.

\subsection{Inter-OS binary differences}

In the previous study~\cite{Scaria2017}, we showed that pre-processing 
pipelines of the Human Connectome Project~\cite{Glasser2013} were 
sensitive to operating system variation so that Figure 
\ref{fig:tissue_class} illustrates the binarized errors after brain 
tissue segmentation process in FreeSurfer. In addition, 
Figure~\ref{fig:fnirt_result} shows the same errors occurred in 
non-linear registration process between CentOS6 and CentOS7 in PreFreeSurfer.

@@

\begin{figure}[H]
\centering
  \includegraphics[scale=0.6]{images/fnirt_result.png} 
  \caption{Absolute errors between FNIRT results from PreFreeSurfer 
  (T2w\_acpc\_to\_MNI\_nonli.nii.gz), subject 104820 (CentOS6 vs. 
  CentOS7)~\cite{Scaria2017}. } 
  \label{fig:fnirt_result}
\end{figure}

@@

\subsubsection{PreFreeSurfer pipeline}

Among the 117 data files produced by 
PreFreeSurfer, 21 did not have any error for any subject, 92 had errors 
for all subjects and 4 had errors for 3 subjects only.

@@

\subsubsection{FreeSurfer pipeline}

In the second analysis, the result of 
PreFreeSurfer is considered as the input files of FreeSurfer pipeline. 
From the FreeSurfer results, total 61 files were found to be different 
between CentOS6 and CentOS7. The processes that introduce errors in 
FreeSurfer are shown on Figure~\ref{fig:fs_error_table}. Most frequent 
processes that create errors are \emph{mris\_anatomical\_stats} which 
measures a number of anatomical properties, \emph{mris\_make\_surface} 
to generate pial surfaces, first pass, no T2 adjustment, 
\emph{mri\_convert} which normalize T1w and T2w images for the benefit 
of \emph{mris\_make\_surfaces}, and \emph{fslmaths} create and apply a 
binary mask to filling of holes to remove  any non-grey/white tissue 
using a threshold created by FSLSTATS at Mean. In addition, all the 
other processes that created errors in the pipeline at least once are 
categorized as others such as \emph{mri\_ca\_register}, 
\emph{mri\_ca\_normalize}, and \emph{mri\_watershed}. Furthermore, 
processes that remove errors in FreeSurfer represented on 
Figure~\ref{fig:fs_remove_table} including \emph{mri\_convert} and 
\emph{mri\_surf2surf}.

@@

\note{Pipeline amplify small numerical differences because they are numerically 
unstable. Furthermore, math libraries evolve over time, leading to 
different numerical errors. we listed some of the irreproducibility 
causes of the pipelines as we mentioned in the previous section 
overally along with exact command arguments.

Mention that this was only possible because the unprocessed data was 
shared in the first place. DICOM to Nifti conversion was out of scope 
and may introduce other issues.}

@@

\note{Our technique is able to characterize the stability of a pipeline's 
components automatically. The numerical instability in the 
PreFreesurfer HCP pipeline arises mainly from linear and non-linear 
registration processes implemented in FSL FLIRT and FNIRT.

@@

The results still suffer from small perturbations literally because of 
the fact that pipeline are not numerically stable. The preferred 
solution is to detect and fix numerical instability of the pipeline 
instead of masking the problem. These processes need to be reviewed to 
understand and correct the cause of instabilities. }
